Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2015-04-06 pituba/article/article_spe.tex
fixing

\begin{abstract}
This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The black-oil equations are integrated in reservoir volume and quantities such as fluid properties are spatially averaged. The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at the datum level. The definition of a convenient datum level is also discussed. It is found that the physical correctness of PVT properties
% obtained from laboratory experiments or empirical correlations
is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
Finally, a numerical scheme for solving the material balance equation is proposed based on nonlinear root finding algorithms.
\end{abstract}

@@

The equation developed by \cite{schilthuis1936active} did not account for the non-existence of hydrodynamic, chemical or thermal equilibrium. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law. To account for the hydrodynamic non-equilibrium, black-oil equations are usually solved for in three dimensional models that aim to predict the pressure field responsible for reservoir fluid flow.

Many works have reported applications of the material balance equation in the literature, but few have explored its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume. This work develops the concept of negligible flow potential and proposes an averaging scheme for flow properties. Subsequently, we suggest how the pressure computed from material balance equation can be compared to the pressure field in 3D black-oil models.

@@

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures (and temperatures) observed during production so that a pseudo-gas component can be defined. Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from oil phase composition. After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other. Since temperature was assumed constant, oil phase density becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:
\begin{equation}
\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
\end{equation}

@@

Spatially averaged quantities are defined for porosity ($\phi$), phase saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha = 0$. This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.

@@

%
%
%
%
%
%Additional assumptions are then needed to simpli
%
%% ERRADO: Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
%
%What is then the meaning of pressure calculations in material balance equations?
%
%For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium. 
%
%We shall see that to compute the averaged phase density, we may define some datum level and use the corresponding local density as an approximation to the averaged density without introducing any important error.

 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase,
\begin{equation}\label{eq: grado}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o \, ,
\end{equation}
 and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}\label{eq: gradodz}
 \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

Assuming reservoir temperature to be sufficiently above gas phase critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

>>>>

\begin{abstract}
This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations. The black-oil equations are integrated in reservoir volume and quantities are spatially averaged. The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure. It is shown that the calculated pressure corresponds to the local pressure at the datum level. The definition of a convenient datum level is also discussed. It is found that the physical correctness of PVT properties
% obtained from laboratory experiments or empirical correlations
is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
Finally, a numerical scheme for solving the material balance equation is proposed based on nonlinear root finding algorithms.
\end{abstract}

@@

The equation developed by \cite{schilthuis1936active} did not account for the non-existence of hydrodynamic, chemical or thermal equilibrium. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law. Black-oil equations are usually solved for in three dimensional models that aim to predict the pressure field responsible for reservoir fluid flow.

Many works have reported applications of the material balance equation in the literature, but few have explored its connection with black-oil equations. \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume. This work develops on those concepts and proposes an averaging scheme for flow properties. Subsequently, it suggest how the pressure computed from material balance equation can be compared to the pressure field in 3D black-oil models.

@@

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressure and temperature observed during production so that a pseudo-gas component can be defined. Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from the oil phase composition. After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other. Since temperature was assumed constant, oil phase density becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in the oil phase:
\begin{equation}
\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
\end{equation}

When the oil phase is undersaturated, the system has an additional degree of freedom and both pressure and the amount of dissolved gas must be known to define its thermodynamic state, e.g., $\rho = \rho(p_o,Y_o)$.

@@

Spatially averaged quantities are defined for porosity ($\phi$), phase saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that pressure at any time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.

@@

%
%
%
%
%
%Additional assumptions are then needed to simpli
%
%% ERRADO: Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
%
%What is then the meaning of pressure calculations in material balance equations?
%
%For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium. 
%
%We shall see that to compute the averaged phase density, we may define some datum level and use the corresponding local density as an approximation to the averaged density without introducing any important error. 
\begin{equation}
\rho_o=\rho_o\left(p_o,Y_o\right)
\end{equation}

\begin{equation}
\rho_o\left( t,z\right) \approx \rho_o\left(t,z_{datum}\right)\left[1+\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} dz\right]
\end{equation}
For an undersaturated oil with no compositional gradient,
 %As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For an undersaturated oil with no compositional gradient,
 \begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)\frac{\partial p_o}{\partial z} = c_o \rho_o g \, ,
 \end{equation}
 where $c_o$ is the isothermal compressibility.

 For a saturated oil,
 \begin{equation}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_T = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_T = 
\frac{1}{\rho_o}\frac{d}{d p_o}\left(\frac{\rho_o^{ST}+R_s\rho_g^{ST}}{B_o}\right) 
 \end{equation}

@@

Assuming reservoir temperature to be sufficiently above gas phase critical temperature, we have a negligible variation of the Z-factor with pressure, $\partial Z / \partial p_g \approx 0$. We use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}
